draft

### 2025-05-26 draft.tex
Updates from Overleaf

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a pretrained deep-learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Compared to engineered features, deep learning approaches may pose a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset, in addition to having access to a GPU. Foundational models aim to lower this barrier, but can fail to generalize well for out-of-distribution data (data which differs sufficiently from the training datasets) \cite{azadFoundationalModelsMedical2023}. Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.  Unlike learned features, whose interpretation requires complex methods like counterfactual techniques or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts, such as the colocalisation of two proteins \cite{garcia-fossaInterpretingImagebasedProfiles2023}.

@@

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
\caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogs. \textbf{A,B,C.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images. \textbf{D.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap.svg}
\caption{\label{fig:astrocytes}\textbf{A,B.} Example pair of astrocyte nuclei image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Bottom panel.} SHAP values of the most important features for classifying the day of differentiation (out of three days). Each point represents a single 3D image containing multiple nuclei, the magnitude of SHAP values indicate how important those features are for the classifier. Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value. Shape features refer to the nucleus and intensity features refer to the DNA staining. The test data accuracy is shown in bold.}
\end{figure}

\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
To demonstrate the broad applicability of cp\_measure, we developed a morphology featurizer with it at its core for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labeled.

The recent proliferation of black-box foundation models trained solely on morphological data \citep{Chen2024-ix, Xu2024-mn} highlights morphology as a highly informative and predictive modality. However, the features these models are typically not interpretable, preventing direct biological assessment. In contrast, using cp\_measure to featurize such images will yield more interpretable features.

However, directly applying CellProfiler to spatial datasets is not convenient since it imposes a nested structure on input data and also cannot handle extremely large images since the full image is loaded into RAM for processing. Since spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screening data, easily 500.000 cells per image, the standalone nature of cp\_measure allowed us to tailor a workflow to such data. For this, Squidpy leverages the powerful data format SpatialData \citep{Marconato2024-SpatialData} in which images are stored as zarr objects. This allows us to parallelize the feature extraction with cp\_measure at the mask level, for example individual cells, streaming batches of cells across computational cores. This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler. Furthermore, this establishes morphological profiling within the scverse ecosystem \citep{Virshup2023-scverse}, exposing it to the wider single-cell community.

To illustrate the value of this workflow, we evaluated its usefulness in identifying morphologically similar regions within two human breast cancer sections profiled with the spatial transcriptomics technology Xenium by 10x Genomics \citep{10x-Genomics2023-el}. We generated a tessellating pattern of hexagonal masks across both images, of which XXXXXX were found to overlay the tissue. These masks were then featurized with the described cp\_measure-based workflow. We first preprocessed the features as described in 
\citep{serranoReproducibleImagebasedProfiling2025}, then integrated both samples using Harmony \citep{Korsunsky2019-Harmony}, and finally clustered them with the Leiden algorithm through Scanpy \citep{wolfSCANPYLargescaleSinglecell2018}.

For each cluster, we extracted the cells within its hexagons and calculated the proportions of cell types predicted by the spatial gene-expression data. A visual comparison of panels A and B in Figure \ref{fig:spatial_omics} shows that morphology-only clustering largely recapitulates the underlying tissue architecture. Panel C further reveals that some clusters have distinct cell-type compositions: for example, cluster 2 is dominated by fibroblasts, whereas adipocytes appear almost exclusively in cluster 5-7.

\begin{figure}[htbp]
\centering
\includegraphics[width=.99\linewidth]{./figs/fig_4_spatial.png}
\caption{\label{fig:spatial_omics}{}Morphology-driven tumor subregion discovery. \textbf{Panel A}: Spatial map of cell types inferred from transcriptomics. \textbf{Panel B}: Morphological subregions obtained by applying Leiden clustering to image-derived feature embeddings.. \textbf{Panel C}: Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type (color legend as shown). Of note is that the cell segmentation in the top-left quarter of the image seems to have performed poorly, since no cells were annotated, whereas the morphological clusters cover these regions.}
\end{figure}

These results demonstrate how cp\_measure can now easily be used outside of its traditional domain. Using it to featurize extremely large images, we show that it produces biologically meaningful morphological clusters that align with - and complement - transcriptomic phenotypes. By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly integrate morphology into their spatial omics analyses, unlocking new insights into tissue organization and cell–cell interactions.

@@

\bibliographystyle{icml2025}
\bibliography{bibliography}
\section{Appendix}
\label{sec:orgdd18dd8}
\subsection{Methods}
\label{sec:orgb3e9382}
\subsubsection{Data and software}
\label{sec:orgbda0ae2}
The code for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}. All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://github.com/afermg/2025\_cpmeasure/}. The datasets we produced for this work are available on Zenodo, and the latest version can be found on \url{https://zenodo.org/records/15390631/latest} and \url{https://zenodo.org/records/15505477/latest}.
\subsubsection{Acknowledgements}
\label{sec:acknowledgements}

>>>>

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a pre-trained deep-learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Compared to engineered features, deep learning approaches may pose a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset, in addition to having access to a GPU. Foundational models aim to lower this barrier, but can fail to generalize well for out-of-distribution data (data which differs sufficiently from the training datasets) \cite{azadFoundationalModelsMedical2023}. Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.  Unlike learned features, whose interpretation requires complex methods like counterfactual techniques or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts, such as the colocalisation of two proteins \cite{garcia-fossaInterpretingImagebasedProfiles2023}.

@@

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
\caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogues. \textbf{Panels A-C.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images. \textbf{Panel D.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/example_shap.svg}
\caption{\label{fig:astrocytes}\textbf{Panels A and B.} Example pair of astrocyte nuclei image and masks. The 3D images were projected over the z-axis, taking the maximum value across the z-stack. \textbf{Panel C.} SHAP values of the most important features for classifying the day of differentiation (out of three days). Each point represents a single 3D image containing multiple nuclei, the magnitude of SHAP values indicate how important those features are for the classifier. Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value. Shape features refer to the nucleus and intensity features refer to the DNA staining. The test data accuracy is shown in bold.}
\end{figure}

\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
To demonstrate the broad applicability of cp\_measure, we developed a morphology featurizer with it at its core for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labelled.

Using CellProfiler on spatial datasets is not convenient due to its enforcement of a nested structure on input data and, since it requires loading entire images into RAM, struggles. Since images from spatial datasets contain thousands of times more cells than conventional microscopy data, usually reaching 500.000 cells per image, the stand-alone nature of cp\_measure allowed us to tailor a workflow to match this data throughput. We used Squidpy to leverage the SpatialData data format \citep{Marconato2024-SpatialData}, enabling us to parallelize the feature extraction with cp\_measure to streaming batches of individual cells across multiple processing cores. This approach enables efficient computation on large-scale datasets, which would be intractable using CellProfiler. Furthermore, this brings morphological profiling into the single-cell -- scverse -- ecosystem \citep{Virshup2023-scverse}, exposing it to the wider single-cell community.

To illustrate the value of this workflow, we used it to identify morphologically similar regions within two human breast cancer sections profiled with the spatial transcriptomics technology Xenium \citep{10x-Genomics2023-el}. We tessellated hexagonal masks across both images, of which XXXXXX were found to overlay the tissue. These masks were then featurized with a cp\_measure-based workflow: We first preprocessed the features following the standard method
\citep{serranoReproducibleImagebasedProfiling2025}, then integrated our two samples using Harmony \citep{Korsunsky2019-Harmony}, and finally obtained the Leiden clusters via Scanpy \citep{wolfSCANPYLargescaleSinglecell2018}.

For each cluster, we extracted the cells within its hexagons and calculated the cell type proportions predicted by gene-expression. A visual comparison of panels A and B in Figure \ref{fig:spatial_omics} shows that morphology-only clustering largely recapitulates the underlying tissue architecture. Panel C reveals that some clusters have distinct cell-type compositions: for example, cluster 2 is dominated by fibroblasts, whereas adipocytes appear almost exclusively in clusters 5-7.

\begin{figure}[htbp]
\centering
\includegraphics[width=.99\linewidth]{./figs/fig_4_spatial.png}
\caption{\label{fig:spatial_omics}{}Morphology reveals the composition of tumour subregions. \textbf{Panel A.} Spatial map of cell types inferred from transcriptomics. \textbf{Panel B.} Morphological subregions obtained by applying Leiden clustering to image-derived feature embeddings. \textbf{Panel C.} Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type (colour legend as shown). The top left quarter of the image shows that the morphological approach worked to predict regions while the expression annotations did not.}
\end{figure}

These results demonstrate how cp\_measure can can be used outside traditional microscopy. When applied on extremely large images it produces biologically meaningful morphological clusters that validate and complement transcriptomic phenotypes. By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly add morphology to their spatial omics workflows, unlocking new insights into tissue organization and cell–cell interactions.

@@

\bibliographystyle{icml2025}
\bibliography{bibliography}
\section{Appendix}
\label{sec:orgdd18dd8}
\subsection{Methods}
\label{sec:orgb3e9382}
\subsubsection{Data and software}
\label{sec:orgbda0ae2}
The codebase for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}. All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://github.com/afermg/2025\_cpmeasure/}. The datasets we produced for this work are available on Zenodo. The selection of samples from JUMP and the images can be found on \url{https://zenodo.org/records/15390631/latest}. The image-based profiles produced with Cellprofiler and used for the comparisons in Figure 2 can be on \url{https://zenodo.org/records/15505477/latest}.

\subsubsection{Acknowledgements}
\label{sec:acknowledgements}
